---
title: "Stat 1500 Group project"
author: "Ian Nyabereka, Ike Nyabereka, Anthony Delaney, Fawwaz Ibrahim"
date: "2025-04-02"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
library(ggplot2)
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
```

```{r load-data}
#opening and reading the data set, adjust file directory

df = read.csv(unz("~/Downloads/adult.zip", "adult.data"), header = FALSE, na.strings = "?", stringsAsFactors = FALSE)
# Convert "?" to NA while reading, Prevent automatic conversion to factors
# Paste the path of the dataset

#Editing column names with the appropriate names 
colnames(df) = c('Age','Work_Class','Fnlwgt','Education','Education_Num','Marital_Status',
                 'Occupation','Relationship_Status','Race','sex','Capital_Gain',
                 'Capital_Loss','Hrs_wrked_per_wk','Native_Country','Income')

#Converting '?' to <NA>
df = as.data.frame(lapply(df, trimws), stringsAsFactors = FALSE) #trimws removes white space in '?' as we suspect there is white space due to difficulty to remove with less obscure methods
df[df == "?"] = NA # Replaces "?"  values with NA

# Percentage of missing values
percentage1 <- sum(is.na(df$Native_Country)) / nrow(df) * 100
percentage2 <- sum(is.na(df$Occupation)) / nrow(df) * 100
percentage3 <- sum(is.na(df$Work_Class)) / nrow(df) * 100
percentage1; percentage2; percentage3

# Remove rows with any <NA> values
df_clean = na.omit(df)
```
```{r convert-data}
str(df_clean)  #Check the structure of the dataframe
df_clean$Income = ifelse(df_clean$Income == ">50K", 1, 0) #  1 == >50K and 0 == <=50K
df_clean$sex = ifelse(df_clean$sex == "Male", 1, 0) #  1 == Male and 0 == Female


#DESCRIPTIVE STATS
#Converting numeric values from characters to integers
df_clean$Age = as.numeric(df_clean$Age)
df_clean$Education_Num = as.numeric(df_clean$Education_Num)
df_clean$Capital_Gain = as.numeric(df_clean$Capital_Gain)
df_clean$Capital_Loss = as.numeric(df_clean$Capital_Loss)
df_clean$Hrs_wrked_per_wk = as.numeric(df_clean$Hrs_wrked_per_wk)

#Getting summary or numeric values
summary(df_clean[,c('Age','Education_Num','Capital_Gain','Capital_Loss','Hrs_wrked_per_wk')])
sd(df_clean$Age)
sd(df_clean$Education_Num)
sd(df_clean$Capital_Gain)
sd(df_clean$Capital_Loss)
sd(df_clean$Hrs_wrked_per_wk)
```
```{r class-imbalance}
# Check class distribution
table(df_clean$sex)
table(df_clean$Education_Num)
```
```{r desc```{r eda-plots}
#PLOTTING GRAPHS
#Bar Chart: Education Level vs Income
ggplot(df_clean, aes(x = Education_Num, fill = as.factor(Income))) +
  geom_bar(position = "fill") +
  labs(title = "Income Distribution by Education Level", x = "Education Level", y = "Proportion of Income Level") +
  theme_minimal()

#Boxplot: Sex vs Education 
ggplot(df_clean, aes(x = factor(sex), y = Education_Num, fill = factor(sex))) +
  geom_boxplot() +
  labs(title = "Education Level by Gender", x = "Sex (0=F, 1=M)", y = "Education Level")

#Bar Plot: Income vs Sex
ggplot(df_clean, aes(x = factor(sex), fill = factor(Income))) +
  geom_bar(position = 'fill') +
  labs(title = 'Income by Sex', x = 'Sex (0=F, 1=M)', y = 'Proportion') +
  scale_fill_manual(values = c("red3", "green4"), labels = c("<=50K", ">50K"))

# Stacked bar plot of Income vs. Sex vs. Education Level
ggplot(df_clean, aes(x = Education_Num, fill = factor(Income))) +
  geom_bar(position = "fill") +
  facet_wrap(~ sex) +  # Facet by Sex (Male/Female)
  labs(title = "Income vs. Sex vs. Education Level", 
       x = "Education Level", 
       y = "Proportion",
       fill = "Income") +
  scale_fill_manual(values = c("red3", "green4"), labels = c("Income <= 50K", "Income > 50K"))
```
```{r logit-model}
# Logistic Regression Model
logit_model <- glm(Income ~ Education_Num + sex, data = df_clean, family = binomial())
# Model summary
summary(logit_model)
# Odds ratios
exp(coef(logit_model))
```
```{r decision-tree}
tree_model <- rpart(Income ~ Education_Num + sex, data = df_clean, method = "class")
rpart.plot(tree_model)
```
```{r model-performance}
# Model performance evaluation
predictions <- ifelse(predict(logit_model, type = "response") > 0.5, 1, 0)
confusionMatrix(factor(predictions), factor(df_clean$Income))
```
```{r random-forest}
rf_model <- randomForest(factor(Income) ~ Education_Num + sex, data = df_clean, importance = TRUE)
varImpPlot(rf_model)
# Create a continuous income variable (if not already present)
df_clean$Income_Continuous <- ifelse(df_clean$Income == 1, 50000 + runif(nrow(df_clean)), 
                                     25000 + runif(nrow(df_clean)))
```
```{r linear-model}
# Linear Regression
lm_model <- lm(Income_Continuous ~ Education_Num + sex, data = df_clean)
summary(lm_model)
```
```{r t-test}
# T-test for income differences by gender
t.test(Income ~ sex, data = df_clean)
```
```{r anova}
# ANOVA for income differences by education level
anova_model <- aov(Income ~ as.factor(Education_Num), data = df_clean)
summary(anova_model)
TukeyHSD(anova_model)
```
